---
title: "HW 02 - jsm2298"
output: github_document
date: "2026-10-04"
---

# Load Packages
```{r}
library(tidyverse)
library(dplyr)
```


# Question 1:
## Load and Clean NYC transit data: retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).
```{r}
nyc_t_df <- read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |> 
  janitor::clean_names() |> 
  select(line, station_name, station_latitude, station_longitude, contains("route"), entry, vending, entrance_type, ada) |> 
  mutate(entry = case_when(
    entry == "YES" ~ TRUE,
    TRUE ~ FALSE
  )) |> 
  pivot_longer(cols = contains("route"),
               names_to = "route_number",
               values_to = "route_name",
               values_transform = list(route_name = as.character))
```
Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

Answer the following questions using these data:

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.
How many stations are ADA compliant?
What proportion of station entrances / exits without vending allow entrance?

Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?
```{r}
# Print total number of A stations:
total_A = nyc_t_df |> 
  filter(route_name == "A") |> 
  select(line, station_name) |> 
  unique() |> 
  summarize(total_A = n())
paste0("We have ", total_A, " stations that serve the A train.")

# Print total number of ADA compliant stations:
total_A_ada = nyc_t_df |> 
  filter(route_name == "A", ada == TRUE) |> 
  select(line, station_name, ada) |> 
  unique() |> 
  summarize(total_A = n())

paste0("We have ", total_A_ada, " ADA compliant stations that serve the A train.")
```


